---
title: "Regression Models Course Project"
author: "Sebastien Plat"
output: pdf_document
---

```{r setup, echo=FALSE,warning=FALSE,message=FALSE}
a4width<- 8.3
a4height<- 11.7
source("..\\Courses projects\\multiplot.R")
require(datasets)
library(UsingR)
library(ggplot2)
library(dplyr)
library(reshape2)
library(knitr)
opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center')
```

# Executive Summary

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

"Is an automatic or manual transmission better for MPG"
"Quantify the MPG difference between automatic and manual transmissions"


# Overview

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

A data frame with 32 observations on 11 variables.

[, 1]	 mpg	 Miles/(US) gallon
[, 2]	 cyl	 Number of cylinders
[, 3]	 disp	 Displacement (cu.in.)
[, 4]	 hp	   Gross horsepower
[, 5]	 drat	 Rear axle ratio
[, 6]	 wt	   Weight (1000 lbs)
[, 7]	 qsec	 1/4 mile time
[, 8]	 vs	   V/S
[, 9]	 am	   Transmission (0 = automatic, 1 = manual)
[,10]	 gear	 Number of forward gears
[,11]	 carb	 Number of carburetors

```{r}
data(mtcars)
str(mtcars)
plot(factor(mtcars$am), mtcars$mpg)
```




